MSDS 7333 Case Study 06

### 2017-06-21 MSDS7333_CaseStudy06_MBaldree_BBrock_TElkins_AKelly.Rmd
Cropped images to reduce white-space, minor wordsmithing, fixed links to reference documents

The case study in chapter 1 of the book *Data Science in R* by Nolan and Lang presented an indoor posiitoning system (IPS) for WiFi devices on a floor of a building at the University of Mannheim. The IPS was trained and tested with data collected at the University with a hand-held device on a grid of 166 points spaced one meter apart in the hallways. The algorithm for the IPS is the K-Nearest Neighbors (k-NN) [1](#REF01).

@@

@@

@@

![Figure 1 - An example of k-Nearest Neighbors from scikit-learn [2](#REF02)](knnExample.png)

@@

The majority of the code in this case study was leveraged from [1](#Ref01). The code was refactored for readibility and modified for part 1 and 2 analysis.

@@

```{r}
plotSignalMaps(c(chosenAP, rejectedAP))
#  figure 2 - number of reports at each location for the rejected and chosen APs.
```

@@

```{r}
df = subset(offline, posX ==2 & posY == 12 & !(mac %in% remainingAP), c(signal, angle, mac))
plotBoxplotSignalStrength(df)
# figure 3 - Signal strength distribution across both Access Points
```

@@

```{r}
# Let's focus on the fixed point (33,3)
df = subset(offline, posX == 33 & posY == 3 & !(mac %in% remainingAP), c(signal, angle, mac))
plotBoxplotSignalStrength(df)
# figure 4 - Signal strength distribution for a point at the other end of the building
```
This plot is similar to the previous distribution but with no outliers. At this point we know signal strength is different for the two access points. We don't know if this will affect our prediction accuracy. Now, let us analyze the two AP's signal strength distribution over angles.

@@

```{r, fig.height=10}
df = subset(offline, posX == 24 & posY == 4 & !(mac %in% remainingAP), c(signal, angle, mac))
plotDensitySignalStrength(df)
# figure 5 - Signal strength distribution across angles
```
The above plot shows the rejectedAP "dd:cd" has a higher signal density at lower strength while the acceptedAP "e1:c0" has a higher density coverage at a higher strength. The distributions show many of the plots are normally distributed while some of them, (e.g., "dd:cd" at 180 and "e1:c0" at 315) are heavily skewed or even bimodal. Looking at each AP from top to bottom, there seems to exist an obvious pattern with respect to angle. This would suggest the median of each distribution varies by angle which can be attributed to a difference in perceived strength. This means the angle of the device contributes to the APs strength of signal.

@@

```{r}
df = subset(offlineSummary, mac %in% c(chosenAP, rejectedAP), c(sdSignal, avgSignal, mac))
plotStdDevSignalStrength(df)
# figure 6 - Standard deviation of average signal strength
```

@@

### 1.2.4 <a id='knn'>K-Nearest Neighbors</a>
k-NN is a non-parametric machine learning algorithm that is instance-based learning as opposed to generalized learning. Our first implementation of k-NN learns based on the number of neighbors within signal strength distance to the point in question from the training points.[2](#Ref02)

@@

Table the test data to review the frequency for each angle and each coordinate. 
```{r}
online = readData("online.final.trace.txt", subMacs = macs)
online$posXY = paste(online$posX, online$posY, sep = "-")
length(unique(online$posXY))
tabonlineXYA = table(online$posXY, online$angle)
#output shows measurements were taken at all angles throughout the floor
tabonlineXYA
# Table 5 - Test data counts per location and angle
```
The above output shows that each angle has test data meaning we can test all angle for accuracy.

Let's Reshape the test data where each AP is in a column. 
```{r}
onlineSummary = castOnline(online)
#confirm rejected AP is included
dim(onlineSummary)
names(onlineSummary)
# Table 6 - Reshaped test data set
```

@@

  print(paste(ap, err))
}
# Table 7 - Accuracies for the various scenarios
```
The above results show that including both APs yielded a better prediction accuracy followed by keeping chosenAP and then rejectedAP. The author's choice to keep chosenAP was indeed better than rejectedAP, but keeping both of them is even better.

@@

  # loop through neighbors
  for (k in 1:K) {
    estFold = predXY(newSignals = testFold[ , 6:12],
                     newAngles = testFold[ , 4],
                     trainFold, numAngles = 3, k = k)
    err[k] = err[k] + calcError(estFold, actFold)
  }
}
# plot k to sum of squared errors
plotSSErrors(err, K)
# figure 7 - estimated ideal 'k'
```
Given the plot above, the best k shows 9, but this changes if you rerun the cross-validation. Experimentation with k yielded 5 with the lowest error of 210. Thus, it was used below.

@@

![Figure 10 - Red lines indicate distance error between predicted location (*) and true location (black squares) using k=4](errors.png)

@@

*Table 8 - Comparison of predicted location errors between power-based and signal-strength-based approaches*

@@

*Table 9 - Nearest neighbors for the target's signal strength*

@@

*Table 10 - Nearest neighbors for the target's signal strength across two APs*

@@

*Table 11 - Nearest neighbors for the target's strongest and weakest signal strength*

@@

## 2.3: Weighted K-Nearest Neighbors {#WtKNN}
In this exercise, we investigated using an alternative k-nearest prediction method using the weighted 
K-nearest neighbors approach.  The calculation of the Euclidean distance error was found to be 1194, whereas the Mean SS error was found to be 210.  The Euclidean distance weighted k-nearest neighbors method had the highest error, while the Mean SS weighted K-nearest neighbors method had the smaller error.

| k |Euclidean distance error|Mean SS error|
|:-:|:-----------------:|:-----------------:|
| |1194|210|

@@

***
# <a id='References'>References</a>
* <a href='REF01'></a> [1] Nolan, Deborah; Data Science in R; 2015; Predicting Location via Indoor Positioning Systems; Chapter 1.

* <a href='REF02'></a> [2] Scikit-learn k-NN; http://scikit-learn.org/stable/modules/neighbors.html#nearest-neighbor-algorithms.

>>>>

The case study in chapter 1 of the book *Data Science in R* by Nolan and Lang presented an indoor posiitoning system (IPS) for WiFi devices on a floor of a building at the University of Mannheim. The IPS was trained and tested with data collected at the University with a hand-held device on a grid of 166 points spaced one meter apart in the hallways. The algorithm for the IPS is the K-Nearest Neighbors (k-NN) [[1]](#REF01).

@@

`Table 1 - Data Element Description for the training data`

@@

`Table 2 - Data Element Description for the Access Point measurements`

@@

![Figure 1 - An example of k-Nearest Neighbors from scikit-learn [[2]](#REF02)](knnExample.png)

@@

The majority of the code in this case study was leveraged from [[1]](#REF01). The code was refactored for readibility and modified for part 1 and 2 analysis.

@@

```{r, fig.height=8}
plotSignalMaps(c(chosenAP, rejectedAP))
#  Figure 2 - Number of reports at each location for the rejected and chosen APs.
```

@@

```{r}
df = subset(offline, posX ==2 & posY == 12 & !(mac %in% remainingAP), c(signal, angle, mac))
plotBoxplotSignalStrength(df)
# Figure 3 - Signal strength distribution across both Access Points
```

@@

```{r}
# Let's focus on the fixed point (33,3)
df = subset(offline, posX == 33 & posY == 3 & !(mac %in% remainingAP), c(signal, angle, mac))
plotBoxplotSignalStrength(df)
# Figure 4 - Signal strength distribution for a point at the other end of the building
```
This plot is similar to the previous distribution but with no outliers. At this point we know signal strength is different for the two access points. We don't know if this will affect our prediction accuracy. Now, let us analyze the two AP's signal strength distribution over angles.

@@

```{r, fig.height=10}
df = subset(offline, posX == 24 & posY == 4 & !(mac %in% remainingAP), c(signal, angle, mac))
plotDensitySignalStrength(df)
# Figure 5 - Signal strength distribution across angles
```
The above plot shows the rejectedAP "dd:cd" has a higher signal density at lower strength while the acceptedAP "e1:c0" has a higher density coverage at a higher strength. The distributions show many of the plots are normally distributed while some of them, (e.g., "dd:cd" at 180 and "e1:c0" at 315) are heavily skewed or even bimodal. Looking at each AP from top to bottom, there seems to exist an obvious pattern with respect to angle. This would suggest the median of each distribution varies by angle which can be attributed to a difference in perceived strength. This means the angle of the device contributes to the APs strength of signal.

@@

```{r}
df = subset(offlineSummary, mac %in% c(chosenAP, rejectedAP), c(sdSignal, avgSignal, mac))
plotStdDevSignalStrength(df)
# Figure 6 - Standard deviation of average signal strength
```

@@

### 1.2.4 <a id='knn'>K-Nearest Neighbors</a>
k-NN is a non-parametric machine learning algorithm that is instance-based learning as opposed to generalized learning. Our first implementation of k-NN learns based on the number of neighbors within signal strength distance to the point in question from the training points.[[2]](#REF02)

@@

Table the test data to review the frequency for each angle and each coordinate. 
```{r}
online = readData("online.final.trace.txt", subMacs = macs)
online$posXY = paste(online$posX, online$posY, sep = "-")
length(unique(online$posXY))
tabonlineXYA = table(online$posXY, online$angle)
#output shows measurements were taken at all angles throughout the floor
tabonlineXYA
# Table 4 - Test data counts per location and angle
```
The above output shows that each angle has test data meaning we can test all angle for accuracy.

Let's Reshape the test data where each AP is in a column. 
```{r}
onlineSummary = castOnline(online)
#confirm rejected AP is included
dim(onlineSummary)
names(onlineSummary)
# Table 5 - Reshaped test data set
```

@@

  print(paste(ap, err))
}
# Table 6 - Accuracies for the various scenarios
```
The above results show that including both APs yielded a better prediction accuracy followed by keeping chosenAP and then rejectedAP. The author's choice to keep chosenAP was indeed better than rejectedAP, but keeping both of them is even better.

@@

  # loop through neighbors
  for (k in 1:K) {
    estFold = predXY(newSignals = testFold[ , 6:12],
                     newAngles = testFold[ , 4],
                     trainFold, numAngles = 3, k = k)
    err[k] = err[k] + calcError(estFold, actFold)
  }
}
# plot k to sum of squared errors
plotSSErrors(err, K)
# Figure 7 - Estimated ideal 'k'
```
Given the plot above, the best k shows 9, but this changes if you rerun the cross-validation. Experimentation with k yielded 5 with the lowest error of 210. Thus, it was used below.

@@

![Figure 10 - Red lines indicate distance error between predicted location (*) and true location (black dots) using k=4](errors.png)

@@

*Table 7 - Comparison of predicted location errors between power-based and signal-strength-based approaches*

@@

*Table 8 - Nearest neighbors for the target's signal strength*

@@

*Table 9 - Nearest neighbors for the target's signal strength across two APs*

@@

*Table 10 - Nearest neighbors for the target's strongest and weakest signal strength*

@@

## 2.3: Weighted K-Nearest Neighbors {#WtKNN}
In this exercise, we investigated using an alternative k-nearest neighbors prediction method using the weighted 
K-nearest neighbors approach.  The calculation of the Euclidean distance error was found to be 1194, whereas the Mean SS error was found to be 210.  The Euclidean distance weighted k-nearest neighbors method had the highest error, while the Mean SS weighted K-nearest neighbors method had the smaller error.

| k |Euclidean distance error|Mean SS error|
|:-:|:-----------------:|:-----------------:|
|5 |1194|210|

@@

***
# <a id='References'>References</a>
* <a id='REF01'></a> [1] Nolan, Deborah; Data Science in R; 2015; Predicting Location via Indoor Positioning Systems; Chapter 1.

* <a id='REF02'></a> [2] Scikit-learn k-NN; http://scikit-learn.org/stable/modules/neighbors.html#nearest-neighbor-algorithms.
